Designing a high-performance boundary element library with OpenCL and Numba

### 2021-04-14 bem.tex
Proof reading

\section{BOUNDARY ELEMENT METHODS WITH BEMPP}
In this section, we provide a brief introduction to boundary element methods and describe the necessary steps for their numerical discretization and solution.

@@

Many practical problems have a significantly more complex structure and can involve block systems of integral equations. Nevertheless, the fundamental structure of what Bempp-cl does is well described by the above problem and will be described in the following.

>>>>

\section{BOUNDARY ELEMENT METHODS WITH BEMPP}
In this section, we provide a brief introduction to boundary element methods (BEM) and describe the necessary steps for their numerical discretization and solution.

@@

Many practical problems have a significantly more complex structure and can involve block systems of integral equations or even coupling with finite element (FEM) codes. Nevertheless, the fundamental structure of what Bempp-cl does is well described using the above simple problem.

### 2021-04-14 bempp_overview.tex
start working through opencl section
remove reference to section, as sections are not numbered
Proof reading

\section{A high level overview of Bempp-cl}

@@

The main computational cost involved in solving a problem using boundary element methods is due to discretising the boundary integral operator on the left-hand side of \cref{eq:bnd_integral} to obtain the matrix $\dmat{A}$: using dense methods, discretising an operator has quadratic complexity in terms of the number of surface triangles (although for larger problems this cost can be reduced through the use of hierarchical matrices or fast multipole methods). Great care is therefore taken to ensure that the operator assembly routines perform as highly as possible.

Once a user has defined in operator using Bempp-cl, the discretization can be computed by calling the \pyth{weak_form} method. Upon calling this method, a regular integrator will be used to assemble all the interactions between non-adjacent elements, and a singular integrator will be used to compute the interactions between adjacent triangles (if the trial and test spaces are defined on different grids, this second integrator is not needed). Depending on the user's preferences, these integrators will internally use computational routines defined using either Numba or OpenCL.

For OpenCL assembly, the code checks additional parameters, such as the default vector length for SIMD operators (e.g., 4 for double precision and 8 for single precision in Intel AVX2, or 1 if a GPU is used), andwhether the discretization should proceed in single or double precision. The OpenCL kernel is then compiled for the underlying compute device using PyOpenCL and executed. If the computational backend is Numba, the call is forwarded to the corresponding Numba discretization routines and executed.

For simple piecewise constant function spaces or other spaces, where the support of each basis function is localized to a single triangle, only one call to the computational routines is necessary. If the support of basis functions is larger than a single triangle, different threads may need to sum into the same global degree of freedom. We discuss this in more detail in \cref{sec:coloring}.

Outside the operator discretization, Numba is used in the following contexts:
\begin{itemize}
  \item Computing the grid topology: this involves iterating through the grid to compute the neighbour relataionships between triangles.
  \item Definition of local-to-global maps for function spaces: again, this requires traversal through the grid and assigning relationships between global and local indices.
  \item Grid functions: a right-hand side function $f$ can be defined as a Python callable. This is just-in-time compiled via Numba and then the product with the corresponding basis functions is integrated in each triangle via numerical quadrature, again via Numba accelerated routines.
  \item Computing sparse matrices.
\end{itemize}

As the cost of each of these processes is in general much smaller than the cost of operator discretization, these can be permformed using Numba without any need to consider the use of OpenCL for potential further speed up.

>>>>

\section{A HIGH-LEVEL OVERVIEW OF BEMPP-CL}

@@

The main computational cost involved in solving a problem using boundary element methods is due to discretising the boundary integral operator on the left-hand side of \cref{eq:bnd_integral} to obtain the matrix $\dmat{A}$: using dense methods, discretising an operator has quadratic complexity in terms of the number of surface triangles. For larger problems this cost can be reduced through the use of approximate hierarchical matrices or fast multipole methods (FMM), where the price of the improved complexity is significantly more involved data structures and additional approximation errors. Bempp-cl provides interfaces to Exafmm \cite{bempp_exafmm} for large problems. Here, we focus on dense discretisation that is natively implemented in Bempp-cl and is suitable for medium sized problems up to a few ten thousand elements, depending on available memory. Great care needs to be taken to ensure that the quadratic complexity operator assembly routines perform as highly as possible.

Once a user has defined an operator using Bempp-cl, the discretization can be computed by calling the \pyth{weak_form} method. Upon calling this method, a regular integrator will be used to assemble all the interactions between non-adjacent elements, and a singular integrator will be used to compute the interactions between adjacent triangles (if the trial and test spaces are defined on different grids, this second integrator is not needed). Depending on the user's preferences, these integrators will internally use computational routines defined using either Numba or OpenCL.

For OpenCL assembly, the code checks additional parameters, such as the default vector length for SIMD operators (e.g., 4 for double precision and 8 for single precision in Intel AVX2, or 1 if a GPU is used), and whether the discretization should proceed in single or double precision. The OpenCL kernel is then compiled for the underlying compute device using PyOpenCL and executed. If the computational backend is Numba, the call is forwarded to the corresponding Numba discretization routines and executed.

For simple piecewise constant function spaces or other spaces, where the support of each basis function is localized to a single triangle, only one call to the computational routines is necessary. If the support of basis functions is larger than a single triangle, different threads may need to sum into the same global degree of freedom.

Outside the operator discretization, Numba is used in the following contexts:
\begin{itemize}
  \item Computing the grid topology: this involves iterating through the grid to compute the neighbour relataionships between triangles.
  \item Definition of local-to-global maps for function spaces: again, this requires traversal through the grid and assigning relationships between global and local indices.
  \item Grid functions: a right-hand side function $f$ can be defined as a Python callable. This is just-in-time compiled via Numba and then the product with the corresponding basis functions is integrated in each triangle via numerical quadrature, again via Numba accelerated routines.
  \item Computing sparse matrix entries, such as for mass matrices that are required to translate between representations of grid functions through basis coefficients or projections, or when we want to evaluate operator products.
\end{itemize}

As the cost of each of these processes is in general much smaller than the cost of operator discretization, these can be performed using Numba without any need to consider the use of OpenCL for potential further speed up.

### 2021-04-14 intro.tex
Proof reading

\chapterinitial{The Bempp boundary element library} (originally named BEM++) started in 2011 as a project to develop an open-source C++ library for the fast solution of boundary integral equations. The original release came with a simple Python wrapper to the C++ library. Over time, more and more functionality was moved into the Python interface, while computationally intensive routines and the main data structures remained in C++.

@@

\subsection{The three language problem}
The original BEM++ was troubled by what is often called the two language problem. It is common for programming language to be either easy for humans to write (e.g., Python) or easy for computers to run and acheive high performance (e.g., C++). It is not common, however, for a language to do both of these. Due to this, it is common in scientific computing libraries to write a library in a fast low-level language such as C++, while providing a user interface in a higher-level language such as Python.

This was the model followed by BEM++, but the problem would perhaps better be described as a three language problem: as well as the Python and C++ code contained in libraries like this, it is common to also include a significant amount of code in a third interfacing language. In the case of BEM++, this third language was Cython, an extension of Python with C data types that can be compiled to include functionality from C++ libraries.

Due to the three languages involved, making changes to the library would often mean having to duplicate changed in three places, with many class structures duplicated in all three languages. This made seemingly simple changes in onorous tasks, and provided a barrier to new members of the community looking to contribute to the open-source project.

@@

In addition to the performance benefits, the library redesign has greatly improved the issues related to the three language problem. Both OpenCL and Numba are use to compile functions. Each function is provided all the data is needs as inputs, so there is no need to duplicate any class structure outside Python. The need for a third interfacing language is removed, as the interfacing to the OpenCL kernels is handled by PyOpenCL.

Boundary element methods are particularly suited to this library model, as the main performance critical task is the computation of discrete operators. Once fast kernels for this have been implemented, the remaining functionality of the library can be written entirely in Python without any great decrease in performance.

@@

We begin this article by giving an overview of the boundary element method, and looking at how Bempp can be used to implement such problems. Following this, we discuss the implementation of boundary element kernels using OpenCL in more detail, and provide a number of performance benchmarks on different compute devices, including CPUs, AMD, Intel, and Nvidia GPUs.

>>>>

\chapterinitial{The Bempp boundary element library} (originally named BEM++) started in 2011 as a project to develop an open-source C++ library for the fast solution of boundary integral equations. The original release came with a simple Python wrapper to the C++ library \cite{bempp_orig}. Over time, more and more functionality was moved into the Python interface, while computationally intensive routines and the main data structures remained in C++.

@@

\subsection{The three language problem}
The original BEM++ was troubled by what is often called the two language problem. It is common for programming language to be either easy for humans to write (e.g., Python) or easy for computers to run and achieve high performance (e.g., C++). It is not common, however, for a language to do both of these. Due to this, it is common in scientific computing libraries to write a library in a fast low-level language such as C++, while providing a user interface in a higher-level language such as Python.

This was the model followed by BEM++, but the problem would perhaps better be described as a three language problem: as well as the Python and C++ code contained in libraries like this, it is common to also include a significant amount of code in a third interfacing language. In the case of BEM++, this third language was first Swig and later Cython, an extension of Python with C data types that can be compiled to include functionality from C++ libraries.

Due to the three languages involved, making changes to the library would often mean having to duplicate changes in three places, with many class structures duplicated in all three languages. This made seemingly simple changes an onerous task, and provided a barrier to new members of the community looking to contribute to the open-source project.

@@

In addition to the performance benefits, the library redesign has greatly improved the issues related to the three language problem. Both OpenCL and Numba are used to compile functions. Each function is provided all the data it needs as inputs, so there is no need to duplicate any class structure outside Python. The need for a third interfacing language is removed, as the interfacing to the OpenCL kernels is handled by PyOpenCL.

Boundary element methods are particularly suited to this library model, as the main performance critical task is the computation of discrete operators. Once fast kernels for this have been implemented, the remaining functionality of the library can be written entirely in Python without any significant decrease in performance.

@@

We begin this article by giving an overview of the boundary element method, and looking at how Bempp can be used to implement such problems. Following this, we discuss the implementation of boundary element kernels using OpenCL in more detail, and provide a number of performance benchmarks on different compute devices, including CPUs, AMD, Intel, and Nvidia GPUs. We conclude with some thoughts on the advantages and disadvantages of OpenCL and Numba.

### 2021-04-14 main.tex
First draft finished.
Proof reading

\input{intro}
\input{bem}
\input{bempp_overview}
\input{opencl}
\input{numba}

@@

\bibitem{opencl}
OpenCL, \url{https://www.khronos.org/opencl/}.
\end{thebibliography}

>>>>

\input{intro}
\input{bem}
\input{bempp_overview}
\input{opencl}
\input{numba}
\input{performance}

@@

\bibitem{opencl}
OpenCL, \url{https://www.khronos.org/opencl/}.

\bibitem{bempp_maxwell}
M. W. Scroggs, T. Betcke, E. Burman, Wojciech \'{S}migaj, and E. van 't Wout, ``Software frameworks for integral equations in electromagnetic scattering based on Cader\'{o}n identities''
{\it Computers \& Mathematics with Applications}, vol. 74, no. 11, pp. 2897--2914, 2017.
{\textsc{doi}: \href{https://doi.org/10.1016/j.camwa.2017.07.049}{10.1016/j.camwa.2017.07.049}.

\bibitem{bempp_orig}
W. \'{S}migaj, T. Betcke, S. Arridge, J. Phillips, and M. Schweiger,
``Solving boundary integral problems with BEM++,''
{\it ADM Transactions on Mathematical Software}, vol. 41, no. 2, Article 6, 2015.
\textsc{doi}: \href{https://doi.org/10.1145/2590830}{10.1145/2590830}

\bibitem{bempp_exafmm}
T. Wang, C. D. Cooper, T. Betcke, and L. A. Barba, ``High-productivity, high-performance workflow for virus-scale electrostatic simulations with Bempp-Exafmm''
{\it arXiv preprint} arXiv:103.01048, 2021.
\end{thebibliography}

### 2021-04-14 opencl.tex
Proof reading

\subsection{SIMD optimized kernels}
When we are running on a CPU and want to take advantage of available SIMD optimizations, we need to make a few modifications to our approach. The corresponding kernel works similarly to described above, but we compute a batch of interactions between one test triangle and $X$ trial triangles, where $X$ is either 4, 8, or 16 (depending on the number of available SIMD lanes). This strategy allows us to optimize almost all floating point operations within a kernel run for SIMD operation. If the number of trial elements is not divisible by 4, 8, or 16, then the few remaining trial elements are assembled with the standard-non vectorized kernel.

@@

Data races in global assembly routines are a problem whenever different elements need to sum into the same global degree of freedom. To solve this we use standard color coding techniques to split up the computations into chunks that access different data regions. To this end, we define two elements as neighbors if they share at least one global degree of freedom. Based on this relationship we run a simple greedy coloring algorithm in the initialization phase of a function space. This information is then later used to run the compute kernels color by color. OpenCL parallelises over test elements and trial elements. Here we therefore have to iterate over the product space of possible color combinations. In Numba we only parallelise over test elements. Here it is sufficient to iterate over all possible colors in the test space.

>>>>

\subsection{SIMD optimized kernels}
When we are running on a CPU and want to take advantage of available SIMD optimizations, we need to make a few modifications to our approach. The corresponding kernel works similarly to what is described above, but we compute a batch of interactions between one test triangle and $X$ trial triangles, where $X$ is either 4, 8, or 16 (depending on the number of available SIMD lanes). This strategy allows us to optimize almost all floating point operations within a kernel run for SIMD operation. If the number of trial elements is not divisible by 4, 8, or 16, then the few remaining trial elements are assembled with the standard-non vectorized kernel.

@@

Data races in global assembly routines are a problem whenever different triangles need to sum into the same global degree of freedom. To solve this we use standard color coding techniques to split up the computations into chunks that access different data regions. To this end, we define two triangles as neighbors if they share at least one global degree of freedom. Based on this relationship we run a simple greedy coloring algorithm in the initialization phase of a function space. This information is then later used to run the compute kernels color by color. OpenCL parallelises over test elements and trial elements. Here we therefore have to iterate over the product space of possible color combinations. In Numba we only parallelise over test elements. Here it is sufficient to iterate over all possible colors in the test space.

### 2021-04-14 performance.tex
Proof reading

In this section we want to provide a number of performance benchmarks. The tests are running on an Intel i9-9980HK with a base clock of 2.4GHz and a burst clock of 5GHz. The CPU supports AVX, AVX2, and AVX-512.
The tests are running on a Dell Precision 7740 Workstation Laptop with 64GB RAM. As GPU we use an Nvidia RTX 3000 GPU. All benchmark tests were performed in Linux. For OpenCL on the GPU we use the Nvidia GPU drivers and as CPU runtime we compare the open-source POCL driver against the Intel CPU OpenCL runtime environment. All timing runs were repeated several times to make sure that the overhead from running the OpenCL and Numba JIT Compiler did not skew the results. Though hardly noticeable by users, for smaller experiments the JIT phase typically takes longer than the actual computation.

@@

Let us start with benchmarking the dense operator assembly. We assemble the matrix $\dmat{A}$ defined by
$$
\dmat{A}_{ij} &= \int_{\Gamma}\psi_i(\bx)\int_{\Gamma}\frac{\phi_j(\by)}{4\pi |\bx -\by|}(\bx)\ds[\by]\ds[\bx]\\
$$
with $\Gamma$ being the unit sphere. For the basis functions $\phi_j$ and test functions $\psi_i$ we compare two cases, piecewise constant functions for both (P0 functions), or nodal, piecewise linear, and globally continuous functions for both (P1 functions). In the P0 case each triangle is associated with just one piecewise constant function. In the P1 case each triangle is associated with $3$ linear basis functions, one for each node of the triangle.

We first compare the OpenCL CPU performance for the Intel OpenCL runtime, and the POCL OpenCL runtime driver. We run the tests in single precision and double precision. The native vector width for both drivers in single precision is $8$, and and double precision $4$, which corresponds to AVX instructions. In Bempp-cl this means that we assemble in vectorized form one test triangle with $8$ trial triangles in single precision, and with $4$ trial triangles in double precision. Within the assembly almost all floating point instructions are manually vectorized to take advantage of this. We should hence see up a factor 2 speed-up between single and double precision assembly.

\begin{figure}
	\center
	\includegraphics[width=6cm]{img/intel_pocl_laplace_comp.pdf}
	\caption{Comparison of the performance of POCL and the Intel OpenCL runtime for the assembly of the Laplace single-layer boundary operator on a grid with 32,768 elements.}
	\label{fig:intel_pocl_laplace_cmp}
\end{figure}

The results for a grid with 32,768 elements are shown in Figure \ref{fig:intel_pocl_laplace_cmp}. We see the expected speed-up between single precision and double precision evaluation. What is interesting to note is the difference between the Intel and the POCL runtime environment. For P0 basis functions the Pocl driver (black bar) significantly outperforms the Intel driver (blue bar) in both, single and double precision. For P1 basis functions the Intel driver gives better performance. The kernel source code is the same in both cases and manual vectorization of floating point operations is performed in the same way. However, in the P1 case significantly more data needs to be copied between a work-item and the global memory since the result of the interaction of a test triangle with a trial triangle is a local $3\times 3$ matrix containing all interactions between the three test basis functions on a triangle and the three trial basis functions.

In Figure \ref{fig:pocl_single_layer} we compare specifically the performance of single precision and double precision evaluation for the POCL driver for various grid sizes in the case of a P0 basis. We can see that the speed-up for larger grid sizes is even slightly faster than just a factor two. The final data point in this graph corresponds to the grid size used for Figure \ref{fig:intel_pocl_laplace_cmp}.

@@

In Bempp-cl we can easily switch between CPU Assembly, GPU Assembly and Numba Assembly. In Figure \ref{fig:cpu_gpu_numba_compare} we present a table for a grid size 2048 elements to compare these modes.
\begin{figure}
\begin{center}
\begin{tabular}{l|c|c}
	        &   single      &    double\\
	        \hline
	 POCL   &   0.05s       &    0.08s\\
	 Numba  &   0.25s       &    0.25s\\
	 GPU    &   0.11s       &    3s\\
\end{tabular}
\end{center}
\caption{Comparison of POCL, Numba and GPU Assembly for a grid with 2048 elements.}
\label{fig:cpu_gpu_numba_compare}
\end{figure}
The speed differences are striking. GPU assembly in single precision is a factor two slower than CPU assembly (and much slower in double precision due to Nvidia hardware restrictions for double precision). Numba is five times slower for single precision and still around three times slower for double precision than POCL. The GPU behaviour is due to the data transfer. We have done numerous experiments with GPU assembly of dense boundary integral operators. While the GPU kernels themselves are extremely fast, data transfer over the bus is destroying performance. But even for medium sized problems we have to transfer the data back to the main memory as GPU RAM is too limited to keep a number of dense matrices of dimension a few ten thousand on the device.

@@

The Numba performance difference is interesting. The main issue, we think, is significantly less usage of AVX vectorized instructions. We have taken care to optimize the Green's function evaluation for vectorized evaluation. But the loops over integration points and other operations auto-vectorize very badly by just looping over all trial triangle for a single test triangle, as we currently do. We could tune our code for better auto-vectorization in Numba. However, this would be work with little benefit as we have highly optimized hand-tuned OpenCL kernels already. We therefore recommend Numba for operator assembly only as a fallback if no OpenCL runtime is available.

@@

Once an integral equation is solved one is usually interested in evaluating the solution not only at the boundary but also at points away from it. Hence, we want to evaluate
$$
f(\bx) = \int_{\Gamma}g(\bx), \by)\phi(\by) \ds[\by]
$$
for many points $\bx$ away from the boundary $\Gamma$. For example, if we want to visualize a solution the points $\bx$ would arise from some plotting grid. Typical of this operation that we usually only want to do it once or a few times at the end of a calculation. Discretising this operation into a dense matrix and then evaluating the dense matrix-vector product is not practical for larger sizes. For very large problems with hundreds of thousands of elements we use Fast Multipole or other accelerated approximate methods. However, for problem sizes of a few ten thousand elements up to around a hundred thousand elements (depending on the problem at hand), direct evaluation of this integral for every value $\bx$ is highly efficient. We won't go into the details of the corresponding OpenCL kernels here but want to show some results that demonstrate the performance of CPU vs GPU. While for the dense assembly of boundary integral operators the bus transfer of the dense matrix limited performance, here we only need to transfer to the device the vector of coefficients of the basis functions for $\phi$ and then back to the host the values at the points $x$.

For this section we used the evaluation of the electric potential operator, defined by
\begin{align*}
\mathcal{E\mathbf{p}}(\bx) &= ik\int_{\Gamma}\mathbf{p}(\by)g(\bx, \by) \ds[\by]\nonumber\\
&-\frac{1}{ik}\nabla_{\bx}\int_{\Gamma}\text{div}\,\mathbf{p}(\by)g(\bx, \by)\ds[\by]
\end{align*}
with $g(\bx, \by) = \frac{e^{ik|\bx - \by|}}{4\pi|\bx - \by}$ the Helmholtz Green's function, and
$k$ the wavenumber of the problem.
The function $\mathbf{p}(\by)\in\mathbb{R}^3$ is a vectorial function, leading to an overall vectorial solution at each point $\bx$. For details about the implementation of electromagnetic problems in Bempp see XXX. Here, we again use a grid with 32,768 elements and so-called (Rao, Wilton, and Glisson) edge based basis functions. We use 50,000 random evaluation points in the exterior of the unit sphere. We set $k=1.0$, though the value is not relevant for the performance.

In Figure \ref{fig:efield_domain_potential} we compare the performance of GPU evaluation with that of the POCL driver. In single-precision the GPU significantly outperforms the CPU, and even in double-precision the Quadro RTX is faster than the 8-core CPU, even though its hardware is not optimised for fast double-precision operations.

@@

With Bempp-cl we have created a Python platform that achieves high-performance through use of modern JIT technologies. Bempp-cl mixes Numba evaluation for less compute intensive linear complexity loops and sparse matrix generation with highly optimized OpenCL kernels for computationally sensitive dense matrix assembly routines. Basing development on Python instead of classical C/C++ makes it very easy to adapt the library and integrate into other applications with Python interfaces. An example is the recent work of integrating Bempp-cl and the Exafmm library for electrostatic virus-scale simulations [XXX]. Strict separation of the computational backend from the interface layer in the library also makes it easy to integrate further backends. We currently support Numba and OpenCL as compute backends. Other technologies could be added easily. OpenCL itself has proved a valuable choice for the purpose of this library. We have all kernels available in CPU and GPU optimized versions and the user can easily move between CPU and GPU based compute devices by a simple parameter change, allowing us to support a wide range of hardware. Here, we have demonstrated Nvidia benchmarks. But it would have made no difference to use AMD or Intel GPUs.

A disadvantage of our approach is that using OpenCL kernels is mixed-language development due to the OpenCL kernels written in C99. Using Numba throughout would be a much more native Python experience. However, while Numba is constantly improving, achieving optimal performance for complex operations is difficult. OpenCL shines here. It makes explicit SIMD operations very easy through dedicated constructs. Moreoever, OpenCL kernels are completely stack based functions, allowing much better optimisations while Numba needs to create every object dynamically, even very small arrays for objects such as coordinate vectors.

Also, one has to consider for this approach the type of operations that are required to be accelerated. For the dense assembly of integral operators we have very simple data structures that can easily be passed to compute kernels. More complex data structures with larger mixture of data movement operations and computations (e.g. tree-based algorithms), are much harder to accelerate since the Python layer imposes limits here on the performance.

@@

The current focus of further developments is on MPI implementations for cluster computing via \verb@mpi4py@. We have also made big steps forward for large problems by creating a black-box FMM interface that is currently implemented for Exafmm, which has allowed us to solve problems with 10 million elements on a single workstation. This shows that Python focused development (with some native routines in lower-level languages) is a scalable model and we are aiming to exploit this scalable further as we move from single workstation computations to large cluster problems.

>>>>

In this section we want to provide a number of performance benchmarks. The tests are running on a Dell Precision 7740 Workstation Laptop with 64 GB Ram. Its CPU is an Intel i9-9980HK with a base clock of 2.4GHz and a burst clock of 5GHz. The CPU supports AVX, AVX2, and AVX-512.
As GPU we use an Nvidia RTX 3000 GPU. All benchmark tests were performed in Linux. For OpenCL on the GPU we use the Nvidia GPU drivers and as CPU runtime we compare the open-source Pocl driver against the Intel CPU OpenCL runtime environment. All timing runs were repeated several times to make sure that the overhead from running the OpenCL and Numba JIT Compiler did not skew the results. Though hardly noticeable by users, for smaller experiments the JIT phase typically takes longer than the actual computation.

@@

Let us start with benchmarking the dense operator assembly. We assemble the matrix $\dmat{A}$ defined by
$$
\dmat{A}_{ij} = \int_{\Gamma}\psi_i(\bx)\int_{\Gamma}\frac{\phi_j(\by)}{4\pi |\bx -\by|}\ds[\by]\ds[\bx]\\
$$
with $\Gamma$ being the unit sphere. For the basis functions $\phi_j$ and test functions $\psi_i$ we compare two cases, piecewise constant functions for both (P0 functions), or nodal, piecewise linear, and globally continuous functions for both (P1 functions). In the P0 case each triangle is associated with just one piecewise constant function. In the P1 case each triangle is associated with $3$ linear basis functions, one for each node of the triangle.

We first compare the OpenCL CPU performance for the Intel OpenCL runtime, and the Pocl OpenCL runtime driver. We run the tests in single precision and double precision. The native vector width for both drivers in single precision is $8$, and in double precision $4$, which corresponds to AVX instructions. In Bempp-cl this means that we assemble in vectorized form one test triangle with $8$ trial triangles in single precision, and with $4$ trial triangles in double precision. Within the assembly almost all floating point instructions are manually vectorized to take advantage of this. We should hence see up a factor 2 speed-up between single and double precision assembly.

\begin{figure}
	\center
	\includegraphics[width=6cm]{img/intel_pocl_laplace_comp.pdf}
	\caption{Comparison of the performance of Pocl and the Intel OpenCL runtime for the assembly of the Laplace single-layer boundary operator on a grid with 32,768 elements.}
	\label{fig:intel_pocl_laplace_cmp}
\end{figure}

The results for a grid with 32,768 elements are shown in Figure \ref{fig:intel_pocl_laplace_cmp}. We see the expected speed-up between single precision and double precision evaluation. What is interesting to note is the difference between the Intel and the Pocl runtime environment. For P0 basis functions the Pocl driver (black bar) significantly outperforms the Intel driver (blue bar) in both, single and double precision. For P1 basis functions the Intel driver gives better performance. The kernel source code is the same in both cases and manual vectorization of floating point operations is performed in the same way. However, in the P1 case significantly more data needs to be copied between a work-item and the global memory since the result of the interaction of a test triangle with a trial triangle is a local $3\times 3$ matrix containing all interactions between the three test basis functions on a triangle and the three trial basis functions.

In Figure \ref{fig:pocl_single_layer} we compare specifically the performance of single precision and double precision evaluation for the Pocl driver for various grid sizes in the case of a P0 basis. We can see that the speed-up for larger grid sizes is even slightly faster than just a factor two. The final data point in this graph corresponds to the grid size used for Figure \ref{fig:intel_pocl_laplace_cmp}.

@@

In Bempp-cl we can easily switch between CPU Assembly, GPU Assembly and Numba Assembly. In Figure \ref{fig:cpu_gpu_numba_compare} we present a table for a grid size 2048 elements to compare these modes.
\begin{figure}
\begin{center}
\begin{tabular}{l|c|c}
	        &   single      &    double\\
	        \hline
	 Pocl   &   0.05s       &    0.08s\\
	 Numba  &   0.25s       &    0.25s\\
	 GPU    &   0.11s       &    3s\\
\end{tabular}
\end{center}
\caption{Comparison of POCL, Numba and GPU Assembly for a grid with 2048 elements.}
\label{fig:cpu_gpu_numba_compare}
\end{figure}
The speed differences are striking. GPU assembly in single precision is a factor two slower than CPU assembly (and much slower in double precision due to Nvidia hardware restrictions for double precision). Numba is five times slower for single precision and still around three times slower for double precision than Pocl. The GPU behaviour is due to the data transfer. We have done numerous experiments with GPU assembly of dense boundary integral operators. While the GPU kernels themselves are extremely fast, data transfer over the bus is destroying performance. But even for medium sized problems we have to transfer the data back to the main memory as GPU RAM is too limited to keep a number of dense matrices of dimension a few ten thousand on the device.

@@

The Numba performance difference is interesting. The main issue, we think, is significantly less usage of AVX vectorized instructions. We have taken care to optimize the Green's function evaluation for auto-vectorized evaluation in Numba. But the loops over integration points and other operations auto-vectorize very badly by just looping over all trial triangle for a single test triangle, as we currently do. We could tune our code for better auto-vectorization in Numba. However, this would be work with little benefit as we have highly optimized hand-tuned OpenCL kernels already. We therefore recommend Numba for operator assembly only as a fallback if no OpenCL runtime is available.

@@

Once an integral equation is solved one is usually interested in evaluating the solution not only at the boundary but also at points away from it. Hence, we want to evaluate
$$
f(\bx) = \int_{\Gamma}g(\bx), \by)\phi(\by) \ds[\by]
$$
for many points $\bx$ away from the boundary $\Gamma$. For example, if we want to visualize a solution the points $\bx$ would arise from some plotting grid. Typical of this operation that we usually only want to do it once or a few times at the end of a calculation. Discretising this operation into a dense matrix and then evaluating the dense matrix-vector product is not practical for larger sizes. For very large problems with hundreds of thousands of elements we use Fast Multipole or other accelerated approximate methods. However, for problem sizes of a few ten thousand elements up to around a hundred thousand elements (depending on the problem at hand), direct evaluation of this integral for every value $\bx$ is highly efficient. We won't go into the details of the corresponding OpenCL kernels here but want to show some results that demonstrate the performance of CPU vs GPU. While for the dense assembly of boundary integral operators the bus transfer of the dense matrix limited performance, here we only need to transfer to the device the vector of coefficients of the basis functions for $\phi$ and then back to the host the values at the points $\bx$.

For this section we use the evaluation of the electric potential operator, defined by
\begin{align*}
\left(\mathcal{E\mathbf{p}}\right)(\bx) &= ik\int_{\Gamma}\mathbf{p}(\by)g(\bx, \by) \ds[\by]\nonumber\\
&-\frac{1}{ik}\nabla_{\bx}\int_{\Gamma}\text{div}\,\mathbf{p}(\by)g(\bx, \by)\ds[\by]
\end{align*}
with $g(\bx, \by) = \frac{e^{ik|\bx - \by|}}{4\pi|\bx - \by|}$ the Helmholtz Green's function, and
$k$ the wavenumber of the problem.
The function $\mathbf{p}(\by)\in\mathbb{R}^3$ is a vectorial function, leading to an overall vectorial solution at each point $\bx$. For details about the implementation of electromagnetic problems in Bempp see \cite{bempp_maxwell}. Here, we again use a grid with 32,768 elements and so-called RWG (Rao, Wilton, and Glisson) edge based basis functions. We use 50,000 random evaluation points in the exterior of the unit sphere. We set $k=1.0$, though the value is not relevant for the performance.

In Figure \ref{fig:efield_domain_potential} we compare the performance of GPU evaluation with that of the Pocl CPU driver. In single-precision the GPU significantly outperforms the CPU, and even in double-precision the Quadro RTX is faster than the 8-core CPU, even though its hardware is not optimised for fast double-precision operations.

@@

With Bempp-cl we have created a Python platform that achieves high-performance through use of modern just-in-time compilation technologies. Bempp-cl mixes Numba evaluation for less compute intensive linear complexity loops and sparse matrix generation with highly optimized OpenCL kernels for computationally sensitive dense matrix assembly routines. Basing development on Python instead of classical C/C++ makes it very easy to adapt the library and integrate into other applications with Python interfaces. An example is the recent work of integrating Bempp-cl and the Exafmm library to build electrostatic virus-scale simulations \cite{bempp_exafmm}. Strict separation of the computational backend from the interface layer in the library also makes it easy to integrate further backends. We currently support Numba and OpenCL as compute backends. Other technologies could be added easily. OpenCL itself has proved a valuable choice for the purpose of this library. We have all kernels available in CPU and GPU optimized versions and the user can easily move between CPU and GPU based compute devices by a simple parameter change, allowing us to support a wide range of hardware. Here, we have demonstrated Nvidia benchmarks. But it would have made no difference to use AMD or Intel GPUs.

A disadvantage of our approach is that using OpenCL kernels is mixed-language development due to the OpenCL kernels written in C99. Using Numba throughout would be a much more native Python experience. However, while Numba is constantly improving, achieving optimal performance for complex operations is difficult. OpenCL shines here. It makes explicit SIMD operations very easy through dedicated constructs. Moreoever, OpenCL kernels are completely stack based functions, allowing much better compiler optimisations while Numba needs to create every object dynamically, even very small arrays for objects such as coordinate vectors.

Also, one has to consider the type of operations that are required to be accelerated. For the dense assembly of integral operators we have very simple data structures that can easily be passed to compute kernels. More complex data structures with larger mixture of data movement operations and computations (e.g. tree-based algorithms), are much harder to accelerate since the Python layer imposes limits here on the performance.

@@

The current focus of further developments is on letting Bempp-cl take advantage of cluster computing by integrating the \verb@mpi4py@ MPI bindings for Python. We have also made big steps forward for large problems by creating a black-box FMM interface that is currently implemented for Exafmm, which has allowed us to solve problems with 10 million elements on a single workstation. This shows that Python focused development (potentially with some native routines in lower-level languages) is a scalable model and we are aiming to exploit this scalability further as we move from single workstation computations to large cluster problems.

\section{Acknowledgements}
The work of Timo Betcke is supported by Engineering and Physical Sciences Research Council Grant EP/V001531/1.
